---
title: "data exploration"
format: html
editor: visual
---

# Data Exploration

The variables I need to look at to answer my questions are DO_percent, Month, and pH

Normal Set Up

```{r}
rm(list = ls()) #clears R's brain
#Load in packages with the functions I need
library(tidyverse)
library(here)
library(ggfortify)
```

Read in the data

```{r}
hlab <- read.csv(here("Data", "HydroLabData.csv"))
#look at the data
glimpse(hlab)
#don't know what's up with the extra columns so I'll just get rid of them 
hlab1 <- select(hlab, -X, -X.1, -X.2, -X.3, -X.4, -X.5, -X.6, -X.7, -X.8, -X.9, -X.10, -X.11 )
glimpse(hlab1)
  
```

Looks like I have 1635 rows and 12 variables. I need to change native_or_chestnut, season, site, and month into factor variables.

## Check for mistakes

```{r}
hlab1$Month <- as.factor(hlab1$Month)
levels(hlab1$Month)
```

Looks good! There are no typos or errors in the Month column.

## Check for outliers

```{r}
# check DO_percent by visualizing the data
hlab1 |>
  ggplot(aes(Month, DO_percent)) + 
  geom_boxplot(outlier.color = "red") + 
  theme_bw()
```

There are some really high values in September, but I know that the hydrolab that was used to get this data was calibrated and acclimatized so I don't feel its necessary to remove them.

```{r}
#check pH by visualizing the data
hlab1 |>
  ggplot(aes(Month, pH)) + 
  geom_boxplot(outlier.color = "red") + 
  theme_bw()
```

There are a few outliers. I'm confident that its due to the actual conditions in the river and not error in the data collection, so there is no reason to get rid of them.

Everything looks good with the data, so I can save the cleaned up data set in my Data folder.

```{r}
write.csv(hlab1, here("Data", "hlab1.csv"), row.names = FALSE)
# now read it in
hlab <- read.csv(here("Data", "hlab1.csv"))
#refactor the data
hlab$Month <- as.factor(hlab$Month)
```

## Exploratory Data Analysis

Looking at DO_percent and Month

```{r}
#visualize the data
ggplot(hlab, aes(x = Month, y = DO_percent)) + 
  geom_boxplot() + 
  theme_bw()
```

```{r}
#Now get some summary statistics
hlab |> 
  group_by(Month) |> 
  summarise(mean.DO = mean(DO_percent),
            sd.DO = sd(DO_percent)) |> 
  View()
```

Looks like there might be a difference between the average DO% in June and in the other months, and a difference between the average DO% in September and the other months

Now look at DO_percent and pH

```{r}
#Visialize the data
ggplot(hlab, aes(x = DO_percent, y = pH)) + 
  geom_point() + 
  theme_bw()
```

Estimate slope and intercept:

Looks like the slope is positive. Difference in pH: 8.4-6.8=1.6

Difference in DO% : 156-0=156

So the slope = 1.6/156 = 0.0103 and I would estimate the intercept to be at about 6.3
