---
title: "Preliminary Data Analysis"
format: html
editor: visual
---

# CAP DAP Preliminary Data Analysis: ANOVA

Explantation

Set up code

```{r}
rm(list = ls())
library(here)
library(tidyverse)
library(ggfortify)
library(gridExtra)
library(multcomp)

# read in the data
hlab <- read.csv(here("data", "hlab.csv"))
```

Make month a factor variable

```{r}
hlab$Month <- as.factor(hlab$Month)
levels(hlab$Month). # take a look at the levels to triple check that there are no errors
```

Make a histogram

```{r}
ggplot(hlab, aes(x = DO_mgl)) +
  geom_histogram(bins = 30) + 
  facet_wrap(~Month) + # look at DO by month
  theme_bw()
```

I think I'm going to switch the variable I use for the ANOVA to use DO_mgl, or dissolved oxygen measured in milligrams per liter because the measurements in percent go up to 150% which isn't really possible. The DO measured in mg/l stays within physically possible ranges.

Make the model

```{r}
anova_mod <- lm(DO_mgl ~ Month, data = hlab) #Month to predict DO
```

Check Assumptions again because I'm using a slightly different variable

```{r}
autoplot(anova_mod)
```

I think its okay?

Run anova

```{r}
anova(anova_mod)
```

**There is some significant difference in the dissolved oxygen between months (F = 120.6, p \<2.2e-16).**

Summary Table

```{r}
summary(anova_mod)
```

Tukey HSD

```{r}
phc1 <- glht(anova_mod, linfct = mcp(Month = "Tukey"))
cld(phc1) # went right to the compact letter display because it's way easier to read
```

Looks like there is a significant difference between July and the other months, and October and the other months. There is no significant difference between June and September.

Replot the data

```{r}
ggplot(hlab, aes(x = Month, y = DO_mgl, fill = Month)) + 
  geom_boxplot() +
  theme_bw() + 
  theme(legend.position = "none", 
        panel.grid = element_blank()) + 
  scale_fill_manual(values = c(July = "purple", 
                               June = "cornflowerblue", 
                               October = "orange", 
                               September = "cornflowerblue" )) + 
  ylab("Dissolved Oxygen (mg/l)")

#Changed colors based on which groups were different from each other
```
