---
title: "Preliminary Data Analysis"
author: "Kate Gardner"
format: html
editor: visual
---

## CAP DAP Preliminary Data Analysis: Linear Regression

This .qmd file has the preliminary linear regression testing for a relationship between dissolved oxygen (`DO_percent`) and pH (`pH`).

Normal code set up

```{r}
rm(list = ls())
library(tidyverse)
library(here)
library(ggfortify)
library(gridExtra)

#read in the data
hlab <- read.csv(here("Data", "hlab.csv"))
```

Since I already visualized and guessed at the results, I'll jump right to building my model

```{r}
hlab_mod <- lm(pH ~ DO_percent, data = hlab)
```

Now check assumptions

```{r}
autoplot(hlab_mod)
```

I think its okay?

Run linear regression test

```{r}
anova(hlab_mod)
```

```{r}
summary(hlab_mod)
```

There is a significant relationship between pH and DO (F = 684, p \<2.2e-16)

Replot the data

```{r}
ggplot(hlab, aes(x = DO_percent, y = pH, color = Native_or_chestnut)) + 
  geom_point() +
  labs(x = "Dissolved Oxygen (%)", y = "pH") + 
  scale_color_manual(values = c(native = "darkgreen", chestnut = "red"))       # showing chestnut and native for funsies
```
